Multiparty Session Type-safe Web Development in TypeScript

### 2020-01-21 PLACES_2020/paper.tex
PLACES 2020: add related works and conclusion

% Example: naughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players. Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied).

Figure \ref{fig:tictactoe} illustrates a possible implementation of the game as an interactive real-time multiplayer web application. Both players are connected to the game server by bidirectional WebSocket connections and interact with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client. SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser. Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.

@@

This MPST methodology has been applied to target applications written in both statically and dynamically typed languages in \cite{Hybrid2016, Python2017}, but do not apply to the event-driven paradigm of interactive web applications. A proposal that caters for web applications in \cite{PureScript2019} targets PureScript to maximise the static guarantees it can provide, but the strong type system make it difficult to extract common code and the functional paradigm is arguably unidiomatic relative to mainstream web development in JavaScript, which limits developer productivity. JavaScript began as the language of the web browser, but a shallow learning curve, an ever-growing ecosystem of libraries and frameworks (such as React.js for writing SPAs) and a successful server-side single-threaded event loop implementation in \textit{Node.js} has propelled it as the de-facto choice for full-stack web application development in a single language.

Whilst the proposal in \cite{Python2017} could be adapted to verify interactive web applications written in JavaScript, the emergence of Microsoft's TypeScript \cite{TypeScript} with a gradual type system motivates the need for a MPST-based workflow for verifying full-stack TypeScript web applications, such that it is flexible with respect to idiomatic practices and frameworks to be useful in industry whilst also leveraging the language features to maximise the static guarantees it is able to provide. Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.

@@

Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets. We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).

\section{The Scribble Framework}
\label{section:scribble}
Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain. This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}.

We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game} and explain the syntax of the language through this example. 
Scribble protocols are organised into \textit{modules}: line 1 declares the name of the module. Line 2 declares a message type to be used as payload types in the global protocol: this would correspond to an \texttt{import \{ Coordinate as Point \} from './Types'} statement in TypeScript, where the developer provides the definition of the \texttt{Coordinate} interface under a \texttt{Types.ts} file.
Line 4 declares the name of the global protocol (\texttt{Game}) and specifies the participants involved in the protocol as \textit{roles}; the protocol body is defined by lines 5-16.
A message passing statement from a sender (e.g. \texttt{P1}) to a receiver (e.g. \texttt{Svr}) require a \textit{label identifier} followed by a list of \textit{payload types} in parentheses (e.g. \texttt{Pos(Point) from P1 to Svr} on line 5).
Semicolons denote continuations for message passing statement in the same way that they express sequential statement execution in programming languages such as Java.

@@

The \texttt{Game} protocol involves three roles: the server (\texttt{Svr}) and two players (\texttt{P1} and \texttt{P2}). First, \texttt{P1} makes a move by sending a coordinate to place the marker on the game board to \texttt{Svr}. Upon receiving the \texttt{Pos} message, \texttt{Svr} will implement the game logic to decide the outcome of \texttt{P1}'s move and communicate with \texttt{P2} accordingly. We express this as a \texttt{choice} statement (lines 6-15), where \texttt{Svr} decides whether \texttt{P1}'s move resulted in a loss for \texttt{P2} (lines 7-8), stalemate (lines 10-11) or an unfinished game (lines 13-15).

If \texttt{P1} performed a winning move (i.e. the latest coordinate resulted in three-in-a-row for \texttt{P1}), then \texttt{Svr} sends a \texttt{Lose} message to \texttt{P2} and a \texttt{Win} message to \texttt{P1} with the coordinates of the latest move. These coordinates are sent to allow the applications running in the browser for both players to update their version of the game state respectively; how they choose to render this on the DOM is up to their implementation, independent of the protocol. Similar message passing actions are sent in the case of a draw, just with updated labels for \texttt{Draw}.

Otherwise, the game has not ended, so \texttt{Svr} updates both players with the coordinates of the latest move by sending an \texttt{Update} message, and the game proceeds to allow \texttt{P2} to make a move. We express this as a recursive call to the \texttt{Game} protocol using a \texttt{do} statement: note that we swap the role parameters \texttt{P2} and \texttt{P1} to express the turn-taking semantics of the game. In short, the global protocol describes \texttt{P1} and \texttt{P2} taking turns until \texttt{Svr} declares a winner/loser or draw.

We leverage the Scribble toolchain to check for protocol well-formedness. This directly corresponds to the multiparty session type theory of verifying valid local type \textit{projections} for all participants of a global type. The implementation of this validation by the toolchain follow from the algorithmic projection procedures defined in \cite{MPST}. We obtain \textit{endpoint protocols} for each role in a well-formed global protocol. An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved. For example, the endpoint protocol for \texttt{P1} will not include the exchange of messages between \texttt{Svr} and \texttt{P2}. An endpoint protocol corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain. We adopt the formalism of EFSMs in \cite{Hybrid2016}, and use the EFSMs as a basis for API generation.

@@

We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encodings to provide linearity guarantees (\sectionref{section:serverlinear}). Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action. We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (listing \ref{lst:svr}).

@@

\begin{figure}[!h]
\begin{lstlisting}[language=JavaScript, title=\texttt{./NoughtsAndCrosses/Game.ts}]
import { Coordinate as Point } from "./Types";
import { Labels } from "./Constants";

@@

The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings. Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM. The Protocol class constructor requires the WebSocket server and the following handlers:

\paragraph{Initial state} When all roles have joined the session, the Protocol class needs to know what to do in the initial state to start communication, be it sending a message to a client or offering branches to a client. For the \textit{Noughts and Crosses} game server, this would be a callback to handle the receipt of the \texttt{Pos(Point)} message from \texttt{P1}.

\paragraph{Receive handlers for each role} The runtime needs to bind a message event listener to the WebSocket between the server and each (client) role, and needs to know how to handle the incoming messages. For each role, the user is required to pass callbacks that encode the receive states for that role. The type alias for such a handler is the \textit{intersection type} \cite{TypeScriptSpec} of all receive states concerning that role. Considering a hypothetical web-service providing addition and square root functionality to a client role \texttt{C}, if \texttt{S1} is a state expecting to receive \texttt{Add(int, int)} from \texttt{C} and \texttt{S4} is a state expecting to receive \texttt{Sqrt(int)} from \texttt{C}, we define \texttt{type Receive_C = S1 \& S4} as the receive handler type alias and the user needs to pass some value of \texttt{Receive_C} when instantiating the session through the Protocol API.

@@

\paragraph{Repeat use} By adopting a callback-oriented design for server-side API generation, channels are not directly accessed by the programmer which makes \textit{reuse} impossible; only the Protocol API is exposed but channel operations are concealed in a private class. Consider the \textit{Noughts and Crosses} server endpoint: the programmer must pass a callback that handles the receipt of a \texttt{Pos(Point)} message and returns either a \texttt{Lose(Point)}, \texttt{Draw(Point)} or \texttt{Update(Point)} message with the appropriate continuation. Upon receiving the \texttt{Pos(Point)} message from a player, our lightweight runtime invokes the callback to get \texttt{Svr}'s choice from the return value, and perform the appropriate send. This is transparent to the programmer and it is impossible for the programmer to send or receive messages more than once, by design.

@@

\paragraph{Virtual DOM} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly. This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=JavaScript, tabsize=4]
render() { return (
	...
	{board.map((row, x) => (
		board.map((col, y) => {
			const SelectPoint = this.props.Pos('click', event => {
				event.preventDefault()
				return { x, y }
			}
			return <SelectPoint><td>.</td></SelectPoint>
		}
	)}
	...	
)}
\end{lstlisting}  
\captionof{lstlisting}{Example fragment of \textit{Noughts and Crosses} \texttt{P1} implementation.}
\label{lst:clientapp}
\end{figure}

We clarify the semantics using the \textit{Noughts and Crosses} example in listing \ref{lst:clientapp}. \texttt{this.props.Pos} is the factory function prop passed from the session runtime. For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.

\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function. Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop.

@@

\dots

The session runtime component is rendered by the programmer and requires the following properties to instantiate the session and execute the EFSM.

\paragraph{Endpoint URL}

\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components.

% State = React component
% Send transitions = model transition factory
% Receive transitions = handled by 
% Protocol component = handle joining and rendering of correct EFSM component

@@

\paragraph{Unused} Our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event. This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states. Consider a hypothetical protocol where, in the current state, sending \texttt{Quit()} to some server role \texttt{S} transitions to the terminal state: \dots

@@

\begin{figure}
\begin{lstlisting}[language=JavaScript, tabsize=4, title=\texttt{./app.ts}]
const handleP1Move: Receive_P1 = (move: Point) => {
	// User-implemented logic
	board.P1(move)
	if (board.won()) { return [Labels.Lose, move, move] }
	else if (board.draw()) { return [Labels.Draw, move, move] }
	else { return [Labels.Draw, move, handleP2Move] } 
}

@@

We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}. In addition to showing the multiparty session type safety achieved by our generated APIs, we also show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.

@@

We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime. We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result. This custom logic is integrated into our 
\texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (listing \ref{lst:svrprotocol}) are defined using

%\dots encapsulate our custom game logic in a class

% Note: no state space explosion, allows for reusability

@@

\paragraph{Implement states} \dots

\paragraph{Register states} \dots

The beauty of the MPST framework is that implementation is decoupled from protocol conformance, so it would be possible to have different front-end implementations for both players; but for the sake of sharing the same game client implementation between both players for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be.

We also illustrate the flexibility of our approach in being interoperable with other libraries and idiomatic JavaScript practices. Our generated APIs provide an implementation of the communication state machine which conforms to the protocol, but \cite{NoughtsAndCrosses} also integrates with the \textit{Redux} state management library to uniformly keep track of the game state across the game client - for simplicity, we save game state into browser local storage.

@@

\paragraph{Endpoint API generation} \dots
% API generation
% - original Python implementation, generating monitors (Rumi's paper)
%	- Gradual type system lets us do better than just dynamic runtime checks
%	- could do the same approach with generating JavaScript monitors, but we have TS and Dart and can leverage more powerful type systems to do more static/compiler checks

% - Java (hybrid approach)
%	- leverage strongly typed system for static behavioural typing
%	- minimse dupliction by defining states in terms of IO interfaces (same IO state with different successor can still share code)
%	- runtime linearity checks

\paragraph{Session types in web development} \dots
% Session types in web development
% - purescript looking at the whole workflow
	% - problem: PureScript isn't as prevalent
% - simon fowler paper on MVU targeting linearity for frontend GUI

\paragraph{Typestate programming} \dots
% Typestates (Mungo)
% - the React components we generate is essentially behaving like typestates of the role, where 
% each role has its set of permitted actions.

\section{Conclusion and Future Works}
We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol.

Future works include \dots

@@

%\section{Introduction}
%% Emergence of interactive JavaScript-intensive applications that maintain continuous stateful
%
%The optional arguments of {\tt $\backslash$documentclass$\{$eptcs$\}$} are
%\begin{itemize}
%\item at most one of
%{\tt adraft},
%{\tt submission} or
%{\tt preliminary},
%\item at most one of {\tt publicdomain} or {\tt copyright},
%\item and optionally {\tt creativecommons},
%  \begin{itemize}
%  \item possibly augmented with
%    \begin{itemize}
%    \item {\tt noderivs}
%    \item or {\tt sharealike},
%    \end{itemize}
%  \item and possibly augmented with {\tt noncommercial}.
%  \end{itemize}
%\end{itemize}
%We use {\tt adraft} rather than {\tt draft} so as not to confuse hyperref.
%The style-file option {\tt submission} is for papers that are
%submitted to {\tt $\backslash$event}, where the value of the latter is
%to be filled in in line 2 of the tex-file. Use {\tt preliminary} only
%for papers that are accepted but not yet published. The final version
%of your paper that is to be uploaded at the EPTCS website should have
%none of these style-file options.
%
%By means of the style-file option
%\href{http://creativecommons.org/about/license/}{creativecommons}
%authors equip their paper with a Creative Commons license that allows
%everyone to copy, distribute, display, and perform their copyrighted
%work and derivative works based upon it, but only if they give credit
%the way you request. By invoking the additional style-file option {\tt
%noderivs} you let others copy, distribute, display, and perform only
%verbatim copies of your work, but not derivative works based upon
%it. Alternatively, the {\tt sharealike} option allows others to
%distribute derivative works only under a license identical to the
%license that governs your work. Finally, you can invoke the option
%{\tt noncommercial} that let others copy, distribute, display, and
%perform your work and derivative works based upon it for
%noncommercial purposes only.
%
%Authors' (multiple) affiliations and emails use the commands
%{\tt $\backslash$institute} and {\tt $\backslash$email}.
%Both are optional.
%Authors should moreover supply
%{\tt $\backslash$titlerunning} and {\tt $\backslash$authorrunning},
%and in case the copyrightholders are not the authors also
%{\tt $\backslash$copyrightholders}.
%As illustrated above, heuristic solutions may be called for to share
%affiliations. Authors may apply their own creativity here.
%
%Exactly 46 lines fit on a page.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill6\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill11\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%
%Here starts a new paragraph. The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill16\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill21\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill26\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill31\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill36\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill41\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.\hfill46\\
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%The rest is like any normal {\LaTeX} article.
%We will spare you the details.
%
%\section{Prefaces}
%
%Volume editors may create prefaces using this very template,
%with {\tt $\backslash$title$\{$Preface$\}$} and {\tt $\backslash$author$\{\}$}.
%
%\section{Bibliography}
%
%We request that you use
%\href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
%{\tt $\backslash$bibliographystyle$\{$eptcs$\}$}
%\cite{bibliographystylewebpage}. Compared to the original {\LaTeX}
%{\tt $\backslash$biblio\-graphystyle$\{$plain$\}$},
%it ignores the field {\tt month}, and uses the extra
%bibtex fields {\tt eid}, {\tt doi}, {\tt ee} and {\tt url}.
%The first is for electronic identifiers (typically the number $n$
%indicating the $n^{\rm th}$ paper in an issue) of papers in electronic
%journals that do not use page numbers. The other three are to refer,
%with life links, to electronic incarnations of the paper.
%
%Almost all publishers use digital object identifiers (DOIs) as a
%persistent way to locate electronic publications. Prefixing the DOI of
%any paper with {\tt http://dx.doi.org/} yields a URI that resolves to the
%current location (URL) of the response page\footnote{Nowadays, papers
%  that are published electronically tend
%  to have a \emph{response page} that lists the title, authors and
%  abstract of the paper, and links to the actual manifestations of
%  the paper (e.g.\ as {\tt dvi}- or {\tt pdf}-file). Sometimes
%  publishers charge money to access the paper itself, but the response
%  page is always freely available.}
%of that paper. When the location of the response page changes (for
%instance through a merge of publishers), the DOI of the paper remains
%the same and (through an update by the publisher) the corresponding
%URI will then resolve to the new location. For that reason a reference
%ought to contain the DOI of a paper, with a life link to corresponding
%URI, rather than a direct reference or link to the current URL of
%publisher's response page. This is the r\^ole of the bibtex field {\tt doi}.
%DOIs of papers can often be found through
%\url{http://www.crossref.org/guestquery};\footnote{For papers that will appear
%  in EPTCS and use \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/eptcs.bst}
%  {\tt $\backslash$bibliographystyle$\{$eptcs$\}$} there is no need to
%  find DOIs on this website, as EPTCS will look them up for you
%  automatically upon submission of a first version of your paper;
%  these DOIs can then be incorporated in the final version, together
%  with the remaining DOIs that need to found at DBLP or publisher's webpages.}
%the second method {\it Search on article title}, only using the {\bf
%surname} of the first-listed author, works best.  
%Other places to find DOIs are DBLP and the response pages for cited
%papers (maintained by their publishers).
%{\bf EPTCS requires the inclusion of a DOI in each cited paper, when available.}
%
%Often an official publication is only available against payment, but
%as a courtesy to readers that do not wish to pay, the authors also
%make the paper available free of charge at a repository such as
%\url{arXiv.org}. In such a case it is recommended to also refer and
%link to the URL of the response page of the paper in such a
%repository.  This can be done using the bibtex fields {\tt ee} or {\tt
%url}, which are treated as synonyms.  These fields should not be used
%to duplicate information that is already provided through the DOI of
%the paper.
%You can find archival-quality URL's for most recently published papers
%in DBLP---they are in the bibtex-field {\tt ee}. In fact, it is often
%useful to check your references against DBLP records anyway, or just find
%them there in the first place.
%
%When using {\LaTeX} rather than {\tt pdflatex} to typeset your paper, by
%default no linebreaking within long URLs is allowed. This leads often
%to very ugly output, that moreover is different from the output
%generated when using {\tt pdflatex}. This problem is repaired when
%invoking \href{http://www.cse.unsw.edu.au/~rvg/EPTCS/breakurl.sty}
%{\tt $\backslash$usepackage$\{$breakurl$\}$}: it allows linebreaking
%within links and yield the same output as obtained by default with
%{\tt pdflatex}. 
%When invoking {\tt pdflatex}, the package {\tt breakurl} is ignored.
%
\nocite{*}
\bibliographystyle{eptcs}
\bibliography{generic}
\end{document}

>>>>

% Example: naughts and crosses
Consider a classic turn-based board game of \textit{Noughts and Crosses} between two players. Both players are identified by either noughts or crosses respectively, and take turns to place a mark on an unoccupied cell of a 3-by-3 grid until one player wins (when their markers form one straight line on the board) or a stalemate is reached (when all cells are occupied). A web-based implementation may involve players connected to a game server via WebSocket connections and interacting with the game from their web browser, which serve a \textit{single-page application} (SPA) of the game client written in a popular framework like \textit{React.js} \cite{React}. SPAs feature a single HTML page and dynamically renders content via JavaScript in the browser. Players take turns to make a move on the game board and the server implements the game logic to progress the game forward until a result (either a win/loss or draw) can be declared.

@@

We see the application of the MPST methodology to verifying interactive web applications written in TypeScript to be an interesting design space - to what extent can the MPST methodology be applied such that we leverage TypeScript language features to maximise static safety guarantees whilst preserving a flexible, natural and idiomatic workflow for TypeScript developers? Such a workflow would ultimately decrease the overhead for incorporating MPST into mainstream web development, which reduces development time by programmatically verifying implementations for communication correctness.

@@

%Figure \ref{fig:workflow} illustrates our proposed development workflow: we extend the Scribble framework (\sectionref{section:scribble}) and generate TypeScript endpoints for server-side (\sectionref{section:server}) and browser-side (\sectionref{section:browser}) targets. We will use the \textit{Noughts and Crosses} game as our running example in the rest of the paper and show how our approach is a practical compromise that combines benefits from static session typing and being compatible with common libraries and idiomatic JavaScript practices (\sectionref{section:example}).

\section{The Scribble Framework}
\label{section:scribble}

Development begins by specifying the permitted communications between participants as a \textit{global protocol} in Scribble, a MPST-based framework introduced in \cite{Scribble} which features a protocol specification language and code generation toolchain. This means that a protocol specified in Scribble maps to some \textit{global type} in multiparty session type theory \cite{MPST}. We specify the \textit{Noughts and Crosses} game as a Scribble protocol in Listing \ref{lst:game}.

@@

We leverage the Scribble toolchain to check for protocol well-formedness. This directly corresponds to the multiparty session type theory of verifying valid local type \textit{projections} for all participants of a global type. The implementation of this validation by the toolchain follow from the algorithmic projection procedures defined in \cite{MPST}. We obtain \textit{endpoint protocols} for each role in a well-formed global protocol. An endpoint protocol only preserves the interactions defined by the global protocol in which the target role is involved, and corresponds to an equivalent \textit{Endpoint Finite State Machine} (EFSM) in the Scribble toolchain. We use the EFSMs as a basis for API generation and adopt the formalisms in \cite{Hybrid2016}.

@@

We make a key design decision \textit{not} to expose communication channels in the TypeScript session type encodings to provide linearity guarantees (\sectionref{section:serverlinear}). Our encodings sufficiently exposes seams for the developer to inject their business logic, whilst the generated session API (\sectionref{section:serversessionapi}) handles the sending and receiving of messages; as a result, the encodings do not concern with the role involved in the send/receive action. We outline the encodings below using examples from the \textit{Noughts and Crosses} game server (Listing \ref{lst:svr}).

@@

\begin{figure}[!h]
\begin{lstlisting}[language=JavaScript]
import { Coordinate as Point } from "./Types";
import { Labels } from "./Constants";

@@

The \textit{Protocol API} is a class that interfaces between the session runtime and the user implementation of the EFSM state encodings. Upon construction, the Protocol class waits for all client roles to connect via WebSocket, then proceeds to the initial state of the EFSM and performs state transitions depending on the permitted communication patterns from the EFSM.

@@

\paragraph{Repeat use} We do not expose channels to the programmer, which makes \textit{reuse} impossible. For example, to send a message, the generated API only requires the payload that needs to be sent, and the session runtime performs the send internally, guaranteeing this action is done \textit{exactly once} by construction.

@@

\paragraph{Virtual DOM (VDOM)} React components are rendered on a logical abstraction of the DOM, which in turn performs a \texttt{diff} on the current browser DOM and patches the delta accordingly. This allows the programmer to directly specify what should be rendered without having to worry about which elements to append or remove from the browser DOM.

@@

\begin{figure}[!h]
\begin{lstlisting}[language=JavaScript, tabsize=4]
render() {
	...
	{board.map((row, x) => (
		board.map((col, y) => {
			const SelectPoint = this.props.Pos('click', (event: UIEvent) => {
				event.preventDefault()
				return { x, y }
			}
			return <SelectPoint><td>.</td></SelectPoint>
		}
	)}
	...	
}
\end{lstlisting}  
\captionof{lstlisting}{Model transition for message sending in 
\textit{Noughts and Crosses} \texttt{P1} implementation.}
\label{lst:clientapp}
\end{figure}

We clarify the semantics using the \textit{Noughts and Crosses} example in Listing \ref{lst:clientapp}. \texttt{this.props.Pos} is the factory function prop passed from the session runtime. For each x-y coordinate on the game board, we create a \texttt{SelectPoint} React component from the factory function (which reads ``build a React component that sends the \texttt{Pos} message with x-y coordinates as payload when the user clicks on it'') and we wrap a table cell (the game board is rendered as a HTML table) inside the \texttt{SelectPoint} component to bind the click event on the table cell.

\subparagraph{Receive} The payload of a message receive transition is passed as a prop to the successor state and can be handled in the successor's view function. Upon receiving the message from the channel, the session runtime renders the successor of the receive state and propagates the payload as the prop accordingly.

@@

Our session runtime can be interpreted as an abstraction on top of the React VDOM that implements the EFSM by construction. The session runtime itself is a React component too, named after the endpoint role identifier: it opens the WebSocket connection to the server, keeps track of the current EFSM state as part of its React component state, and most importantly, renders the React component encoding of the active EFSM state. Channel communications are managed by the runtime, which allows it to render the successor of a receive state upon receiving a message from the channel. Similarly, the session runtime is responsible for passing the required props for model transitions to EFSM state React components. The session runtime component is rendered by the programmer and requires the following props:

\paragraph{Endpoint URL} The session runtime needs this in order to open the WebSocket connection.

\paragraph{Concrete state components} The programmer writes their own implementation of each state (mainly to customise how the state is rendered and inject business logic into state transitions) by extending the abstract React class components. The session runtime requires references to these concrete components order to render the user implementation accordingly.

@@

\paragraph{Unused} Our approach \textit{does not} statically detect whether all transitions in a certain state are bound to some UI event. This means that it is possible for an implementation to \textit{not} handle transitions to a terminal state but still type-check, so we cannot prevent unused states.

@@

\begin{figure}
\begin{lstlisting}[language=JavaScript, tabsize=4]
const handleP1Move: Receive_P1 = (move: Point) => {
	// User-implemented logic
	board.P1(move)
	if (board.won()) { return [Labels.Lose, move, move] }
	else if (board.draw()) { return [Labels.Draw, move, move] }
	else { return [Labels.Draw, move, handleP2Move] } 
}

@@

We apply our framework to implement a web-based implementation of the \textit{Noughts and Crosses} running example in TypeScript; the interested reader can find the full implementation in \cite{NoughtsAndCrosses}. In addition to MPST-safety, we show that our library design welcomes idiomatic JavaScript practices in the user implementation and is interoperable with common front- and back-end frameworks.

@@

We set up the WebSocket server inside an Express.js \cite{ExpressJS} application on top of a Node.js \cite{NodeJS} runtime. We define our own game logic in a \texttt{Board} class to keep track of the game state and expose methods to query the result. This custom logic is integrated into our \texttt{Receive\_P1} and \texttt{Receive\_P2} handlers (Listing \ref{lst:svrprotocol}), so the session runtime can handle \texttt{Pos(Point)} messages from players and transition according to the injected game logic. Note that, by the structural typing nature of TypeScript, replacing \texttt{handleP2Move} on line 6 with \texttt{handleP1Move} would be type-correct - this allows for better code reuse as supposed to defining additionl abstractions to work around the limitations of nominal typing in \cite{Hybrid2016}. There is also full type erasure when transpiling to JavaScript to run the server code, so state space explosions is never a runtime consideration.

@@

We implement the game client for players as a React application - extending from the generated abstract React components, registering our custom implementation as props for the session runtime component. For the sake of code reuse for demonstration purposes, \cite{NoughtsAndCrosses} uses \textit{higher-order components} (HOC) to build the correct state implementations depending on which player the user chooses to be. We leverage the \textit{Redux} state management library to keep track of game state, thus showing the flexibility of our library design in being interoperable with other libraries and idiomatic JavaScript practices.

@@

\paragraph{Endpoint API generation} \cite{Python2017} targeted Python applications and the generation of runtime monitors to dynamically verify communication patterns. Whilst the same approach could be applied to JavaScript, we can provide more static guarantees with TypeScript's gradual typing system and compiler. \cite{Hybrid2016} targeted Java applications and proposed to encode the EFSM states and transitions as classes and instance methods respectively, with behavioural typing achieved statically by the type system and channel linearity guarantees achieved dynamically since channel resources are exposed and aliasing is not monitored.

\paragraph{Session types in web development} \cite{PureScript2019} targeted web development in PureScript using the \textit{Concur UI} framework and proposed a type-level encoding of EFSMs as multi-parameter type classes. However, it presents a trade-off between achieving static linearity guarantees from the type-level EFSM encoding under the expressive type system and providing an intuitive development experience to programmers, especially given the prevalence of JavaScript and TypeScript applications in industry. \cite{MVU2019} focused on applying binary session types in front-end web development and presented approaches that tackle the challenge of guaranteeing linearity in the event-driven environment.

Our work applies the aforementioned approaches in a \textit{multiparty} context in a manner that is compatible with industrial tools and practices to ultimately invite the notion of MPST-safety into interactive web application development in industry.

\section{Conclusion and Future Works}
We have presented a MPST-based framework for developing full-stack interactive TypeScript applications with WebSocket communications that conform to a protocol, statically providing linear channel usage guarantees and affine channel usage guarantees for back-end and front-end targets respectively.

Future works include incorporating \textit{explicit connection actions} introduced in \cite{ExplicitConnections} in our API generation to better model real-world communication protocols that may feature in interactive web applications. Whilst our approach does support multiparty sessions, protocols are required to have exactly one role running on the server and client roles must interact with each other via the server to respect the properties of WebSocket connections, which is the transport abstraction we support. Extending support to WebRTC would cater for peer-to-peer communication between browsers, which further opens up possibilities for communication protocols supported by our approach.

@@

\nocite{*}
\bibliographystyle{eptcs}
\bibliography{generic}
\end{document}
